\section{The Abstract Blockchain Interface}\label{sec:interface}

The testcases in \cref{fig:testcases} are clients that
execute against an abstract blockchain interface
of transaction requests and responses (see the left side of \cref{fig:architecture}).
That interface consists of the methods in \cref{fig:api}, that allow one to use the blockchain
as a persistent storage for object-oriented programming.
%
\begin{figure*}
  \begin{description}
  \item[\texttt{reference = installJar(key, payer, gasLimit, gasPrice, classpath, bytes)}]
  \item[\texttt{reference = callConstructor(key, payer, gasLimit, gasPrice, classpath, constructor[, args...])}]
  \item[\texttt{result = callMethod(key, payer, gasLimit, gasPrice, classpath, method, receiver[, args...])}]
  \item[\texttt{result = runMethod(caller, gasLimit, classpath, method, receiver[, args...])}]
  \end{description}
  \caption{The abstract blockchain interface (ABI) that allows clients to execute
  object-oriented Java code in blockchain.}
  \label{fig:api}
\end{figure*}

Method \texttt{installJar} installs a jar in blockchain.
That jar will typically contain the code of new classes (hence also of new contracts)
that can be later referenced by using \texttt{reference} as classpath for future constructor or method
invocations. The installation is paid by an account \texttt{payer}, whose private \texttt{key} is used
to sign the transaction request. The request is run only if the current gas price is at most
\texttt{gasPrice} and is allowed to consume at most \texttt{gasLimit} units of gas. The \texttt{bytes}
are those of the jar to install. Classes inside such jar can reference other classes, previously
installed in blockchain, through the optional \texttt{classpath} reference. In most cases, the latter
is the reference to the runtime of Takamaka, that has been installed at
\texttt{takamakaLib} during blockchain initialization. The result of \texttt{installJar} is
an abstract, machine-independent \texttt{reference} to where the jar has been installed in blockchain.
For instance, in method \texttt{beforeAll}
in \cref{fig:testcases}, the jar containing the compiled Java bytecode of the contract in
\cref{fig:ponzi} is installed in blockchain at a reference that is saved as \texttt{jar}.
The faucet account of the testnet is used to pay for the installation. The runtime of Takamaka
is used as classpath, so that for instance the library classes \texttt{Contract} and
\texttt{StorageLinkedList} used in \cref{fig:ponzi} get correctly resolved.

Method \texttt{callConstructor} calls a constructor of a class and yields the
machine-independent \texttt{reference} to where the resulting object
has been allocated in blockchain. Its parameters are the same as for \texttt{installJar}, but it needs the
signature of the \texttt{constructor} to call (class name and argument types), instead of the \texttt{bytes}
of a jar, and it needs as many arguments as there are parameters for that constructor. Such arguments can
either be primitive values or machine-independent references to objects that have been already allocated
in blockchain earlier. An example of use of \texttt{callConstructor} is in the two testcases in
\cref{fig:testcases}, to instantiate the contract in \cref{fig:ponzi}, by calling its no-args constructor.
Moreover, \texttt{callConstructor} is also used inside the implementation (not shown) of the
macro \texttt{setAccounts} used in \texttt{beforeEach} in \cref{fig:testcases}, to instantiate new accounts.

Method \texttt{callMethod} calls a method of an already existing object (called \texttt{receiver}) and yields
the optional result of the method (for non-\texttt{void} methods). It requires the same parameters
as \texttt{callConstructor}, but it needs a method signature (class name, method name and argument types)
instead of a constructor signature. Moreover, it requires the machine-independent reference
to the \texttt{receiver}, and as many arguments as there are parameters for the method.
For instance, \texttt{callMethod} is used in the two testcases in \cref{fig:testcases} to call the
\texttt{void} method \texttt{invest} on an already existing \texttt{contract} of the \texttt{Ponzi} class
in \cref{fig:ponzi}. The only argument for \texttt{invest} is the fee to invest.

Method \texttt{runMethod} is similar to \texttt{callMethod} but can only be used to call
side-effect-free methods (also called \emph{view} methods in Solidity terminology). Such methods do not
modify the state database (\cref{fig:architecture}) and consequently needn't be included in blocks and
can be execute \emph{for free}, without requiring a signature nor a payment of gas. Therefore, such calls
have a \texttt{caller} rather than a \texttt{payer}. There is still a maximal
limit to the gas cost of the method, in order to avoid infinite executions. The Hotmoka runtime will
ensure that the executed method has no side-effects, and the transaction will fail otherwise.
For instance, \texttt{runMethod} is used in the implementation of \texttt{getBalanceOf} in \cref{fig:testcases}
(not shown), to call the side-effect-free \texttt{balance} method of a contract.

The ABI in \cref{fig:api} has been explicitly simplified. The interested reader will find
the actual, complete ABI as interface \texttt{io.hotmoka.node.api.Node} in~\cite{hotmokacode}.
The latter has additional methods for calling Java static methods (that do not have a \texttt{receiver}),
for accessing the current state of an object, its index (history),
the manifest and the configuration of the blockchain and the set of responses/requests executed so far.
Moreover, paid methods (as the first three in \cref{fig:api}) require a \emph{nonce} to avoid
request replaying (as in Solidity). We do not show this here for simplicity. In practice, such nonces are automatically
computed by the superclass \texttt{HotmokaTest} in \cref{fig:testcases}. Finally, the ABI in
\cref{fig:api} is synchronous, while the actual ABI has also asynchronous versions of the first three
methods in \cref{fig:api}, that do not yield a reference or result but a future to them. Asynchronicity
allows one sometimes to write faster testcases, when the outcome of a transaction needn't be committed before
a subsequent transaction can be executed. We do not show this here.
