\section{Stubbing the Abstract Blockchain Interface}\label{sec:stubbing}

The testcases in \cref{fig:testcases} can be executed over the abstract
blockchain interface of \cref{sec:interface}, whose implementation can
be an actual blockchain testnet. As seen in \cref{sec:takamaka},
one needs a testnet in order to exploit the
existence of a faucet, useful to establish the fixture of the testcases.
This solution works perfectly and has the advantage of testing the full
execution stack in \cref{fig:architecture}: not just the client executing
the testcases but also the peer, the Hotmoka application, the miners and even
other peers that might be connected through gossiping and might contribute
to the execution of the transaction requests. Namely, \cref{sec:experiments}
will report experiments with the execution of testcases over actual
testnets. But this solution has a clear drawback: the execution time
of the testcases becomes relatively long, since an actual testnet
creates an actual blockchain and the block creation rate cannot be
too small, or otherwise synchronization between the peers might be lost.
With a block creation rate of $10$ seconds (hence quite short) the
execution of the testcases in \cref{fig:testcases} requires around one minute,
since the calls to \texttt{installJar}, \texttt{setAccounts},
\texttt{callConstructor} and \texttt{callMethod} trigger a transaction request
that must be included in the next block before the next call can be executed.
The asynchronous versions of such calls would not help here: these testcases
need to wait until requests get executed and transactions have modified
the state, before checking assertions on the modified state or sending
a new request from the modified state.

Slow testcases means that they cannot be used for continuous integration,
at least not on a regular basis, after each commit. For instance,
they cannot be used as non-regression tests during a modification of
the code of the application in \cref{fig:architecture}. But this was one
of the reasons of this article: allowing improvements and optimizations to
the Hotmoka application and using the testcases on the smart contracts
as non-regression tests during continuous integration.
The solution has been to provide a stubbed version of the blockchain peer,
that just passes calls to the API in \cref{fig:api}
over to the Hotmoka application (\cref{fig:stubbed}).
The rest of the abstract blockchain interface described
in \cref{sec:interface} has been mocked (for instance,
calls about the configuration or the
index of the node). The stubbed peer does not use any database of blocks since
it does not even keep a blockchain. Therefore, there is no block
creation rate and the calls in \cref{fig:api} are executed immediately,
at the maximal speed allowed by the Hotmoka application.
There is no mining nor gossiping, nor other connected peers to synchronize with.

\begin{figure*}[t]
  \begin{center}
    \begin{tikzpicture}
      \draw [fill=verylightviolet,thick] (-0.5,-0.8) rectangle +(4,2.3);
      \draw (1.5,-0.1) node {stubbed peer};

      \draw[->, thick] (-4,-0.45) -- (-0.5,-0.45);
      \draw (-2.25,-0.7) node {transaction requests};
      \draw (-2.25,-0.2) node {{\small API}};
      \draw[<-, thick] (-4,0) -- (-0.5,0);
      \draw (-2.25,0.3) node {transaction responses};
      \draw [fill=verylightblue,thick] (-6,-0.6) rectangle +(2,0.8);
      \draw (-5,-0.2) node {client(s)};

      \draw [fill=verylightgreen,thick] (0.5,2.6) rectangle +(2,1.5);
      \draw (1.5,3.6) node {Hotmoka};
      \draw (1.5,3.2) node {application};
      \draw[-, thick] (0.5,3.4) -- (-1,3.4);
      \draw (-1,3.4) node[scale=0.7] {\db};
      \draw (-1,2.5) node {database of states};

      \draw[<-, thick] (1.2,1.5) -- (1.2,2.6);
      \draw[<-, thick] (1.8,2.6) -- (1.8,1.5);

      \draw[-, dashed] (1.2,1.5) -- (-0.5,0);
      \draw[-, dashed] (1.8,1.5) -- (-0.5,-0.45);
    \end{tikzpicture}
  \end{center}
  \caption{A stubbed blockchain peer.
    It just forwards the abstract blockchain API calls
    from the clients to the Hotmoka application.}
  \label{fig:stubbed}
\end{figure*}

Still, the stubbed peer in \cref{fig:stubbed} has the same abstract blockchain interface
as a real peer. Therefore, testcases such as those in \cref{fig:testcases} can be run
against it, without any modification: a switch in the test execution invocation
provides different instantiations for the peer. This is an application of the
traditional open-closed principle of software engineering~\citep{Meyer97}.

\Cref{sec:experiments} will confirm that testcases executed against a stubbed peer
are quite fast, enough to use them as non-regression tests during continuous
integration, for the development of the Hotmoka application.
