\section{Testing Takamaka in a Nutshell}\label{sec:takamaka}

Takamaka is a minimal, deterministic subset of Java meant for writing Solidity-like
smart contracts. Java was not devised for writing smart contracts, therefore
Takamaka adds specific features that are needed for such contracts but are not
available, normally, in Java: the ability to identify the caller of a method
or constructor (as \texttt{caller()}) and to limit such caller to specific classes
(through the annotation \texttt{@FromContract(class)}) and the ability to transfer
cryptocurrency at a method or constructor call (through the \texttt{@Payable}
annotation). Moreover, the runtime of Takamaka takes care, automatically, of
the serialization and deserialization of objects into the state database
(\cref{fig:architecture}) and it automatically charges and limits the execution
of Takamaka code through the consumption of gas, exactly as in Solidity.
This is achieved through a mixture of code instrumentation and code verification
and it allows the dynamic installation and execution of Java smart contracts in a public
and permissionless blockchain.

This article is not the place for a complete and detailed description of Takamaka,
for which we refer to already published literature~\citep{Spoto19,OlivieriST25}.
The short description in the previous paragraph
should be enough to understand the examples in this paper.

\begin{figure*}
  \begin{lstlisting}[language=Takamaka,numbers=left]
public class Ponzi extends Contract {
 private final StorageList<PayableContract> investors = new StorageLinkedList<>();|\label{line:investors}|

 public @FromContract(PayableContract.class) Ponzi() {
  investors.add((PayableContract) caller());|\label{line:add_constructor}|
 }

 public @Payable @FromContract(PayableContract.class) void invest(long fee) {
  require(fee >= 1000, "you must invest at least 1000");|\label{line:require}|
  var split = fee / investors.size();
  investors.forEach(investor -> investor.receive(split));|\label{line:receive}|
  investors.add((PayableContract) caller());|\label{line:add_invest}|
 }
}
  \end{lstlisting}
  \caption{A Takamaka smart contract that implements a Ponzi scheme.}
  \label{fig:ponzi}
\end{figure*}

Consider the Takamaka smart contract in \cref{fig:ponzi},
translated from an equivalent Solidity smart contract
by~\cite{IyerD18}. It implements a Ponzi scheme
that investors can join either by installing an instance of
the contract (by calling its constructor)
and becoming its first investor (\cref{line:add_constructor}) or,
subsequently, by calling the \texttt{invest} method of an existing instance
of the contract and paying a fee (\cref{line:add_invest}).
In the latter case, the fee is split among all previous investors (\cref{line:receive}).
As long as new investors keep joining, previous investors keep earning split fees.
The last investor will, instead, lose its fee.

Both the constructor of \texttt{Ponzi} and method \texttt{invest} in \cref{fig:ponzi}
are limited to be called from a class that must extend \texttt{PayableContract}, which is
a Takamaka library class that implements contracts able to receive a payment through the
\texttt{receive} method. This is important since investors are kept in a list in the
persisted state of the object (\cref{line:investors}) and must be able to receive
split fees whenever a new investor arrives (\cref{line:receive}). Accounts
are a subclass of \texttt{PayableContract} in Takamaka, controlled through a key pair,
hence accounts can call both the constructor and \texttt{invest} in \cref{fig:ponzi}.
\texttt{StorageList}s (\cref{line:investors}) are a deterministic replacement
of Java's native lists: the latter are not deterministic and therefore cannot be used in blockchain
and are consequently rejected in Takamaka. Finally, note that method \texttt{receive} of
\texttt{PayableContract} is final in Takamaka (\ie it cannot be redefined),
which makes reentrancy impossible at \cref{line:receive} and avoids the complex use of
Solidity's withdrawing pattern~\citep{AntonopoulosWPMP25}, which is needed instead
in the Solidity code in~\cite{IyerD18}.

\begin{figure*}
\begin{lstlisting}[language=Java]
class Ponzi extends HotmokaTest {
 var PONZI = classNamed("io.hotmoka.examples.ponzi.Ponzi");
 var PONZI_CONSTR = ConstructorSignatures.of(PONZI);
 var INVEST = MethodSignatures.of(PONZI, "invest", LONG);
 var GAS_PRICE = 1;
 var GAS_LIMIT = 10_000_000;
 var jar;

 @BeforeAll static void beforeAll() {
  // send to Hotmoka a request to install a jar
  jar = installJar(privateKeyOfFaucet(), faucet(), GAS_LIMIT, GAS_PRICE, takamakaLib(),
                   Files.readAllBytes("ponzi.jar"));
 }

 @BeforeEach void beforeEach() {
   // a macro inherited from HotmokaTest: send to Hotmoka requests to create and fund three accounts:
   // the faucet will pay
   setAccounts(1_000_000_000, 1_000_000_000, 1_000_000_000);
 }

 @Test void feeSplitsCorrectly() {
  // send to Hotmoka a request to let account(0) create the contract and become the first investor
  var contract = callConstructor(privateKey(0), account(0), GAS_LIMIT, GAS_PRICE, jar, PONZI_CONSTR);
  // take note of the balance of account(0) now
  var initial = getBalanceOf(account(0));
  // account(1) invests in the contract
  var first_fee = 10_000_000;
  callMethod(privateKey(1), account(1), GAS_LIMIT, GAS_PRICE, jar, INVEST, contract, first_fee);
  // the first investment went to account(0)
  assertEquals(initial + first_fee, getBalanceOf(account(0)));
  // account(2) invests in the contract
  var second_fee = 2_000_000;
  callMethod(privateKey(2), account(2), GAS_LIMIT, GAS_PRICE, jar, INVEST, contract, second_fee);
  // half of the second investment went to account(0)
  assertEquals(initial + first_fee + second_fee/2, getBalanceOf(account(0)));
 }

 @Test void tooSmallFeeThrowsException() {
  // send to Hotmoka a request to let account(0) create the contract and become the first investor
  var contract = callConstructor(privateKey(0), account(0), GAS_LIMIT, GAS_PRICE, jar, PONZI_CONSTR);
  // account(1) invests too little in the contract
  var fee = 100;
  throwsTransactionExceptionWithCauseAndMessage
   ("io.takamaka.code.lang.RequirementViolationException", "you must invest at least 1000",
    () -> callMethod(privateKey(1), account(1), GAS_LIMIT, GAS_PRICE, jar, INVEST, contract, fee)
   );
 }
}
\end{lstlisting}
\caption{Two Java JUnit testcases for the smart contract in \cref{fig:ponzi}.}
\label{fig:testcases}  
\end{figure*}

\Cref{fig:testcases} shows a JUnit class with two testcases for the smart contract
in \cref{fig:ponzi}. This class is a client of the Hotmoka blockchain, that uses its
ABI (\cref{fig:architecture})
to install the smart contract in \cref{fig:ponzi} and to call its constructor and method.
Namely, in JUnit, method \texttt{beforeAll} is executed only once, before running the
testcases of the class. In our case, it installs in blockchain the compiled Java bytecode
(a Java \emph{jar}) of the contract in \cref{fig:ponzi}. It lets the \emph{faucet}
pay for this transaction. A faucet is an account of a testnet that can be used to
pay for transactions, \emph{for free}, which is useful during testing.
A mainnet, of course, has no faucet account. An abstract reference to the installed
jar is later accessible as \texttt{jar}. Method \texttt{beforeEach}, instead, is
run before \emph{each} of the testcases. It calls the inherited method \texttt{setAccounts}
that executes transactions to create and fund three account objects, controlled by three fresh key pairs.
That macro pays such transactions and funds from the faucet of the blockchain.
An abstract reference to such objects is later accessible by method \texttt{account} and their fresh
private keys through method \texttt{privateKey}. One will need such keys in order to sign transaction
requests to execute the constructor and method in \cref{fig:ponzi}.

The two testcases run in the fixture established by \texttt{beforeAll} and \texttt{beforeEach}.
The first testcase verifies that the fee is split correctly among previous investors.
It initially creates an instance of the contract, by calling its constructor. It makes account $0$
pay for that, hence the transaction request must be signed with the private key of that account.
It writes down the balance of account $0$ and lets account $1$ call \texttt{invest} and become
an investor. It verifies that the full fee went to the only previous investor,
which was account $0$. Then it lets account $2$ call \texttt{invest} and become an investor and
verifies that, this time, only half of the fee went to account $0$.

The second testcase verifies the expected behavior if an investor invests too little
in the contract. Namely, there is a lower bound to the acceptable fee
(\cref{line:require} in \cref{fig:ponzi}). Lower fees lead to a transaction failure because
of a missed requirement. To verify that behavior, the second testcase in \cref{fig:testcases}
creates another instance
of the smart contract and lets account $1$ invest too little in it. It verifies that
an exception is thrown, whose cause has the expected exception class and message.

The testcases in \cref{fig:testcases} set a minimal gas price for the transactions that they execute,
since they are meant to run on a testnet that is configured to ignore gas prices. This avoids
testcase failures due to fluctuating gas prices. An alternative would be to query the current gas price of
the blockchain and use a safe overapproximation, but we do not show it here for simplicity.

If the testcases in \cref{fig:testcases} are run, they both succeed, giving confidence
in the expected behavior of the smart contract in \cref{fig:ponzi}. From another perspective,
they support our confidence in the correct implementation of the Hotmoka application
and peer  (\cref{fig:architecture}) and can be used as
non-regression tests when the code of such application or peer must be modified.
